---
title: "AvianMonitoring-eBird"
author: "Wancheng Sun"
date: "10/27/2018 "
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#install those packages before run the rmd if there is no package following codes:
#install.packages(c("rio","data.table","dplyr","anytime","knitr","kableExtra", "ggplot2","ISOweek","alluvial","lubridate"))

```

## Data Preparation

>Reshaped data of Avia Monitoring

 We reshape some data format and create some new columns which are easy and convient for the following analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE}

#Avia data
library(rio)
avia <- import("C:\\Users\\wsun\\Desktop\\ebird\\Rscript\\Avia.csv") 

library(data.table)
avia <- setDT(avia)[,Date := as.Date(Date, "%m/%d/%Y")]

namevariable <- names(avia)
namevariable <- gsub(" ","_",namevariable)
namevariable <- gsub("/","_",namevariable)
names(avia) <- namevariable

avia2 <- melt(avia[,c(1:3,6:7)], id.vars = c("Monitoring_Route","Date","eBird","H_S"))
avia2 <- setDT(avia2)[!is.na(value),]

avia3 <- melt(avia[,c(1:3,8:9)], id.vars = c("Monitoring_Route","Date","eBird","H_S"))
avia3 <- setDT(avia3)[!is.na(value),]

avia4 <- melt(avia[,c(1:3,10:11)], id.vars = c("Monitoring_Route","Date","eBird","H_S"))
avia4 <- setDT(avia4)[!is.na(value),]

avia5 <- melt(avia[,c(1:3,12:13)], id.vars = c("Monitoring_Route","Date","eBird","H_S"))
avia5 <- setDT(avia5)[!is.na(value),]

avia6 <- melt(avia[,c(1:3,14:15)], id.vars = c("Monitoring_Route","Date","eBird","H_S"))
avia6 <- setDT(avia6)[!is.na(value),]

avia1 <- rbind(avia2,avia3,avia4,avia5,avia6)
rm(avia2,avia3,avia4,avia5,avia6)

avia1 <- setDT(avia1)[,.(value = sum(value)), by = 
                          c("Monitoring_Route","Date","eBird","H_S","variable")]

avia2 <- setDT(avia)[,.(eachCount = sum(`#_of_Species`), totalCount = sum(Total_species)),by = 
                         c("Monitoring_Route","Date","eBird")]
library(dplyr)
aviaFinal <- left_join(avia2,avia1,by = c("Monitoring_Route","Date","eBird"))
rm(avia1,avia2)

library(ISOweek)

aviaFinal <- setDT(aviaFinal)[,Weeks := strftime(Date, format = "%Y-W%V")][,c("year", "WeekNoYear") := tstrsplit(Weeks, "-", fixed=TRUE)]
aviaFinal <- aviaFinal[,firstWeekDay := paste(Weeks,1,sep = "-")][,firstWeekDay := 
                                               ISOweek2date(firstWeekDay)][,firstWeekDay := 
                                                                        strftime(firstWeekDay, format = "%Y-%m-%d")]

library(lubridate)
aviaFinal <- setDT(aviaFinal)[,`:=`(Year= year(Date),Month = month(Date),MonthWithYear = strftime(Date, format = "%Y-%m"))]

aviaFinal <- aviaFinal[,season := ifelse(Month %in% c(3,4,5),"spring",ifelse(Month %in% c(6,7,8),"summer",ifelse(Month %in% c(9,10,11),"autumn","winter")))]

aviaFinal <- aviaFinal[, H_S :=  recode(H_S,"1"= "Heard","2"="Seen","3"="H&S")]

aviaNoYear <- aviaFinal[,.(value = sum(value,na.rm = T)), by= c("WeekNoYear","variable")][!is.na(variable),]

library(knitr)
library(kableExtra)

aviaFinal <- aviaFinal[,-c("year","firstWeekDay")]
kable(sample_n(aviaFinal, 5)) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

```



> Data from eBird

  With the limited access to eBird dataset, we cannot compare this example data with our data directly, we just take this part of example data from eBird website as reference, whcih will make the following analysis easier.


```{r echo=FALSE, message=FALSE, warning=FALSE}

#ebird data

ebird <- import("C:\\Users\\wsun\\Desktop\\ebird\\Rscript\\ebirdWebData.xlsx") 

namevariable <- names(ebird)
namevariable <- gsub("# ","",namevariable)
namevariable <- gsub(" ","_",namevariable)
names(ebird) <- namevariable

ebird <- setDT(ebird)[,Week_starting_on := strftime(Week_starting_on, 
                                                    format = "%m-%d")][,variable := species][,variable := recode(variable,"Wood Duck" = "WODU","Marsh Wren" = "MAWR", "Great Blue Heron" = "GBHE", "Wood Thrush" = "WOTH", "Canada Goose" = "CAGO")]

kable(sample_n(ebird, 5)) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

```

## Descriptive Analysis

```{r echo=FALSE, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}

aviaAllu <- aviaFinal[!is.na(aviaFinal$variable),c(1,6,11,14,7:8)]
aviaAllu <- setDT(aviaAllu)[,.(value = sum(value)),by = c("Monitoring_Route","H_S","variable","Year","season")]
aviaAllu <- aviaAllu[,c("variable","Year","H_S","Monitoring_Route","season","value")]


```




### Cross Table

>  With season

   First, we compare these five speices seasonally which includes autumn, spring, summer and winter.

```{r echo=FALSE, message=FALSE, warning=FALSE}
aviaAllus <- data.table::dcast(aviaAllu,variable ~ season,fun = sum, value.var=c("value"))

kable(head(aviaAllus)) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

> With Monitoring Route

  Secondly, we compare thesee five species based on Monitoring_Route which includes Blue Trail, Green Trail, Purpule Trail and Red Trail.

```{r echo=FALSE, message=FALSE, warning=FALSE}
aviaAllus <- data.table::dcast(aviaAllu,variable ~ Monitoring_Route,fun = sum, value.var=c("value"))

kable(head(aviaAllus)) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

> With H_S

  Thirdly, we comare these five species based on H_S which means the species both heard and seen, headred, seen seperately. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
aviaAllus <- data.table::dcast(aviaAllu,variable ~ H_S,fun = sum, value.var=c("value"))

kable(head(aviaAllus)) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

> With Year

  Forthly, we coompare these five species yearly which mainly focues on 2016, 2017, 2018 years.

```{r echo=FALSE, message=FALSE, warning=FALSE}
aviaAllus <- data.table::dcast(aviaAllu,variable ~ Year,fun = sum, value.var=c("value"))

kable(head(aviaAllus)) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


## ANOVA Analysis

According to descriptive part, We can be sure each variable with each species in the sample is not homogeneous. But five species have their unique characters on each variable, and we may eplore more relationship on the variables.

The total number of observation in MAWR, WOTH and WODU is not much enough to well estimate the variables difference in the sample. We will only take a trying to test anova group defference on the two species(CAGO, GBHE).

### Canada Goose(CAGO)


```{r message=FALSE, warning=FALSE}
#Randomized Block Design
fit <- aov(value ~ H_S + Year + Monitoring_Route + season, aviaAllu[variable == "CAGO",])
anova(fit)

```

As the result shows, we can not reject the null hypothesis on 5%, GAGO's observation number in the four variables has no significant different.


### Great blue hero(GBHE)

```{r message=FALSE, warning=FALSE}
#Randomized Block Design
fit <- aov(value ~ H_S +  season + Year + Monitoring_Route, aviaAllu[variable == "GBHE",])
anova(fit)
```

Still no variable's p-value low than 5% in GBHE. But as the result of the fit and the alluvial diagram suggest, the independence variables have interaction.


```{r message=FALSE, warning=FALSE}

# Two Way Factorial Design 
fit <- aov(value ~  season * H_S, aviaAllu[variable == "GBHE",])
anova(fit)
```

With the involving of the interaction, the H_S variable seems to be significant this time. It means different season has different observation ways number.









## Reference

[Create Awesome HTML Table with knitr::kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)

[Meteorological season](https://en.wikipedia.org/wiki/Season)

[Creating Alluvial Diagrams](https://cran.r-project.org/web/packages/alluvial/vignettes/alluvial.html)

[Factorial Treatment Structure](https://stat.ethz.ch/~meier/teaching/anova/factorial-treatment-structure.html)

[Two way - between subject analysis of variance](http://personality-project.org/r/r.guide/r.anova.html)

[Bird Observations - eBird](https://ebird.org/barchart?byr=2016&eyr=2018&bmo=1&emo=12&r=US-OH)

[ggplot2 Cheatsheet](https://ggplot2.tidyverse.org)


